\textbf{ Analytic Light Curves in Reflected Light: \\% Phase Curves, Occultations, and Non-Lambertian Scattering for Spherical Planets }

### 2020-05-18 tex/ms.tex

% Title
\title{%
    \textbf{
        Analytic Light Curves in Reflected Light:%
        \\%
        Phase Curves, Occultations, and Non-Lambertian Scattering
    }
}

@@

\begin{abstract}
    We derive efficient, closed form, differentiable, and numerically stable
    solutions for the flux measured from a planet or moon seen in reflected
    light, either in our out of occultation.
    %
    Our expressions apply to the computation of scattered
    light phase curves of exoplanets, secondary eclipse light curves in
    the optical, or future measurements of planet-moon and planet-planet
    occultations, as well as to photometry of solar system bodies.
    %
    We derive our solutions for Lambertian bodies illuminated by a point
    source, but extend them to model illumination sources of finite
    angular size and rough surfaces with phase-dependent scattering.
    %
    Our algorithm is implemented in \Python within the open-source
    \starry mapping framework and is designed with efficient gradient-based
    inference in mind.
    %
    The algorithm is ${\sim}4-5$ orders of magnitude faster than direct
    numerical evaluation methods and ${\sim}10$ orders of magnitude more
    precise.
    %
    We show how the techniques developed here may one day lead to the
    construction of two-dimensional maps of terrestrial planet surfaces,
    potentially enabling the detection of continents and oceans on
    exoplanets in the habitable zone.
    %
    \href{https://github.com/rodluger/starrynight}{\color{linkcolor}\faGithub}
\end{abstract}

@@

There is a vast literature on techniques for mapping exoplanet surfaces
based on their phase curves
\citep[e.g.,][]{Russell1906,Lacis1972,Knutson2007,Cowan2008,Oakley2009,Berdyugina2017}
and occultation light curves
\citep[e.g.,][]{Williams2006,Rauscher2007,Majeau2012,deWit2012,Rauscher2018},
both in thermal and reflected (scattered) light.
In particular, much attention has been given to techniques for mapping
Earth-like planets from visible-light reflected phase curves
\citep[e.g.,][]{Ford2001,Kawahara2010,Kawahara2011,Fujii2012,Kawahara2020,Aizawa2020}.
Unlike thermal phase curves, which primarily encode (often degenerate)
information about longitudinal surface brightness variations
\citep{Russell1906}, reflected light curves often contain information
about the full two-dimensional surface albedo distribution
\citep[e.g.,][]{Kawahara2010}.

@@

Many studies have analyzed real Earth reflected light curves to
infer surface properties of our planet as an exercise in
prepartion for the mapping of exoplanets.
%
\citet{Cowan2009} analyzed visible-light disk-integrated light curves
of the Earth taken by the Deep Impact spacecraft to produce
longitudinal maps of the surface, harnessing multi-band observations
to disentangle static surface brightness features from temporally
variable clouds. More recently, \citet{Jiang2018} and \citet{Fan2019}
used data from the L1-stationed DSCOVR satellite to infer surface
and cloud properties of the Earth, and \citet{Luger2019b} analyzed
background scattered light in TESS photometry to reconstruct a cloud
map of the Earth.

@@

The present paper is an extension to the \starry algorithm,
adapting it to model phase curves and occultation light curves
in reflected light. The expressions
we derive are analytic: they may all be expressed in closed form
in terms of algebraic operations involving trigonometric functions
and (at times) elliptic integrals. We derive numerically
stable recursion relations for the efficient evaluation of all
expressions and code them within an autodifferentiation framework
to enable the computation of high precision derivatives for use in
gradient-based inference and optimization schemes.
Our code is fully
\href{https://github.com/rodluger/starry}{open-source},
comprehensively
\href{https://dev.azure.com/rodluger/starry/_test/analytics?definitionId=4}{unit-tested},
and supplemmented with an extensive
\href{https://rodluger.github.io/starry}{API documentation}
and suite of
\href{https://rodluger.github.io/starry/latest/tutorials.html}{tutorials}.
%
As in \citet{Luger2019}, in the caption of each of the figures
we provide links \codeicon\ to the exact \Python scripts that
generated them. Next to many of the equations we also provide
links \prooficon\ to \textsf{Jupyter} notebooks containing
detailed derivations and/or validations.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_i \neq \vartheta_r$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.

@@

Whereas in \citet{Luger2019} we compute the observed flux by always
integrating over the unocculted portion of the disk, here we find that
it is often easier and more computationally efficient to compute the
integral of the intensity over the
\emph{simplest} region -- meaning the one with the fewest boundaries --
and combine it with the formalism from \S\ref{sec:solution-no-occ} to
compute the visible flux. These integrals may be over the unocculted dayside,
the occulted dayside, the unocculted nightside, or the occulted nightside;
in the case of the latter three, a bit of algebra
(\S\ref{sec:cases-pathological}--\S\ref{sec:cases-pathological}) is needed
to relate these to the observed flux. Additionally, in some cases we can
avoid computing new integrals entirely, as the solution can be obtained from
a combination of the classical \starry solution vector $\sT$ and the
formalism from the unocculted case (\S\ref{sec:solution-no-occ}).

@@

Cases with zero roots (case 1 -- case 4) are treated in \S\ref{sec:cases-easy},
while cases with one or two roots (case 5 -- case 8) are treated in
\S\ref{sec:cases-hard}. Cases with three or four roots
(case 9 -- case 10) are rarely encountered
in practice, but are possible for some pathological configurations; these are
treated in \S\ref{sec:cases-pathological}.

@@

\subsection{The term $\mathbb{p}_2$}
\label{sec:p2}
%
The final integral we must solve is that corresponding to
$\mathbb{p}_2$ ($\mu = \nu = 1$). As in \citet{Luger2019}, this is
the integral of the linear limb darkening term,
whose solution must be handled
separately due to the fact that the corresponding antiderivative
in Equation~(\ref{eq:G}) is not a polynomial in $x$, $y$, and $z(x, y)$.
The integral we must solve is
%
\begin{align}
    \label{eq:p2}
    \mathbb{p}_2 & =
    \vint{\pmb{\phi}}{
        \frac{1}{3}
        \left(
        \frac{
            1 - z(r_o\cos\varphi, b_o + r_o\sin\varphi)^3
        }{
            1 - z(r_o\cos\varphi, b_o + r_o\sin\varphi)^2
        }
        \right)
        \left(r_o^2 + b_o r_o \sin\varphi\right)
    }
    \quad,
\end{align}
%
where $z$ is the usual Cartesian coordinate (Equation~\ref{eq:z}).
The solution is tricky, but fortunately a similar integral was solved in
Equation~(34) of \citet{Pal2012}. Adapting their solution to our formalism,
we obtain
%
\begin{proof}{p2}
    \label{eq:p2_soln}
    \mathbb{p}_2 & =
    \frac{1}{3}
    \Big(
    c_0 +
    c_1 \DF +
    c_2 \DE +
    c_3 \DPi
    \Big)
\end{proof}
%
where
%
\begin{proof}{p2}
    c_0 &=
    \Delta
    \Bigg\{
    -\atantwo\left(
    -(b_o - r_o) \cos\valpha, (b_o + r_o) \sin\valpha
    \right)
    + \valpha
    \nonumber \\
    &\qquad\enspace\enspace
    - \frac{4}{3} b_o r_o
    \sin\valpha \cos\valpha
    \sqrt{1 - (b_o - r_o)^2 - 4 b_o r_o \sin^2\alpha}
    \nonumber \\
    &\qquad\enspace\enspace
    + \pmb{\delta}(b_o, r_o, \valpha)
    \Bigg\}
    \nonumber \\[0.5em]
    c_1 &=
    \frac{1 + b_o^4 - b_o^2(5 + 2 r_o^2) + r_o^4 + r_o^2}{3 \sqrt{1 - (b_o - r_o)^2}}
    \nonumber \\[0.5em]
    c_2 &= \frac{(b_o^2+7r_o^2-4)\sqrt{1 - (b_o - r_o)^2}}{3}
    \nonumber \\[0.5em]
    c_3 &= \frac{b_o + r_o}{(b_o - r_o)\sqrt{1 - (b_o - r_o)^2}}
    \quad,
    \\
    \intertext{and}
    %
    \delta(b_o, r_o, \alpha) &=
    \begin{cases}
        -2\pi
         &
        \qquad
        \alpha > \frac{3\pi}{2} \,\, \text{and} \,\, b_o > r_o
        \\
        +2\pi
         &
        \qquad
        \alpha > \frac{3\pi}{2} \,\, \text{and} \,\, b_o < r_o
        \\
        0
         &
        \qquad
        \text{otherwise}
        \quad.
    \end{cases}
\end{proof}
%
The quantities $\pmb{F}(\valpha \,|\, \nicefrac{1}{k^2})$ and
$\pmb{E}(\valpha \,|\, \nicefrac{1}{k^2})$
are the same incomplete elliptic integrals as those in
\S\ref{sec:J}, while
%
\begin{align}
    \label{eq:Pi}
    \Pi(n; \psi \,|\, m) & \equiv
    \int_0^{\psi}
    \frac{\dd \varphi}{(1 - n \sin^2\varphi)\sqrt{1 - m \sin^2 \varphi}}
\end{align}
%
is the incomplete elliptic integral of the third
kind, with
%
\begin{proof}{p2}
    \label{eq:n}
    n & = -\frac{4 b_o r_o}{(r_o - b_o)^2}
    \quad.
\end{proof}
%

>>>>

% Title
\title{%
    \textbf{
        Analytic Light Curves in Reflected Light:%
        \\%
        Phase Curves, Occultations, and Non-Lambertian Scattering for Spherical Planets
    }
}

@@

\begin{abstract}
    We derive efficient, closed form, differentiable, and numerically stable
    solutions for the flux measured from a spherical planet or moon seen in reflected
    light, either in our out of occultation.
    %
    Our expressions apply to the computation of scattered
    light phase curves of exoplanets, secondary eclipse light curves in
    the optical, or future measurements of planet-moon and planet-planet
    occultations, as well as to photometry of solar system bodies.
    %
    We derive our solutions for Lambertian bodies illuminated by a point
    source, but extend them to model illumination sources of finite
    angular size and rough surfaces with phase-dependent scattering.
    %
    Our algorithm is implemented in \Python within the open-source
    \starry mapping framework and is designed with efficient gradient-based
    inference in mind.
    %
    The algorithm is ${\sim}4-5$ orders of magnitude faster than direct
    numerical evaluation methods and ${\sim}10$ orders of magnitude more
    precise.
    %
    We show how the techniques developed here may one day lead to the
    construction of two-dimensional maps of terrestrial planet surfaces,
    potentially enabling the detection of continents and oceans on
    exoplanets in the habitable zone.
    %
    \href{https://github.com/rodluger/starrynight}{\color{linkcolor}\faGithub}
\end{abstract}

@@

There is an extensive literature on techniques for mapping spherical exoplanet surfaces
based on their phase curves
\citep[e.g.,][]{Russell1906,Lacis1972,Knutson2007,Cowan2008,Oakley2009,Berdyugina2017}
and occultation light curves
\citep[e.g.,][]{Williams2006,Rauscher2007,Majeau2012,deWit2012,Rauscher2018},
both in thermal and reflected (scattered) light.
In particular, much attention has been given to techniques for mapping
Earth-like planets from visible-light reflected phase curves
\citep[e.g.,][]{Ford2001,Kawahara2010,Kawahara2011,Fujii2012,Kawahara2020,Aizawa2020}.
Unlike thermal phase curves, which primarily encode (often degenerate)
information about longitudinal surface brightness variations
\citep{Russell1906}, reflected light curves often contain information
about the full two-dimensional surface albedo distribution
\citep[e.g.,][]{Kawahara2010}.

@@

Many studies have analyzed real Earth reflected light curves to
infer surface properties of our planet as an exercise in
prepartion for the mapping of exoplanets.
%
\citet{Cowan2009} and \citet{Cowan2011b} analyzed visible-light disk-integrated light curves
of the Earth taken by the Deep Impact spacecraft to produce
longitudinal maps of the surface, harnessing multi-band observations
to disentangle static surface brightness features from temporally
variable clouds.  A transit of Earth by the Moon was observed
as well \citep{Livengood2011}, although this data has yet to be
exploited for mapping purposes.
More recently, \citet{Jiang2018} and \citet{Fan2019}
used data from the L1-stationed DSCOVR satellite to infer surface
and cloud properties of the Earth, and \citet{Luger2019b} analyzed
background scattered light in TESS photometry to reconstruct a cloud
map of the Earth.

@@

The present paper is an extension to the \starry algorithm,
adapting it to model phase curves and occultation light curves
in reflected light. The expressions
we derive are analytic: they may all be expressed in closed form
in terms of algebraic operations involving trigonometric functions
and (at times) elliptic integrals. We derive numerically
stable recursion relations for the efficient evaluation of all
expressions and code them within an autodifferentiation framework
to enable the computation of high precision derivatives for use in
gradient-based inference and optimization schemes.
Our code is fully
\href{https://github.com/rodluger/starry}{open-source},
comprehensively
\href{https://dev.azure.com/rodluger/starry/_test/analytics?definitionId=4}{unit-tested},
and supplemented with an extensive
\href{https://rodluger.github.io/starry}{API documentation}
and suite of
\href{https://rodluger.github.io/starry/latest/tutorials.html}{tutorials}.
%
As in \citet{Luger2019}, in the caption of each of the figures
we provide links \codeicon\ to the exact \Python scripts that
generated them. Next to many of the equations we also provide
links \prooficon\ to \textsf{Jupyter} notebooks containing
detailed derivations and/or validations.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_i \neq \vartheta_r$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.  Finally, the presence
of time-variable features, such as clouds, dust storms, or seasonal
variations of vegetation are not accounted for in the model.

@@

Whereas in \citet{Luger2019} we compute the observed flux by always
integrating over the unocculted portion of the disk, here we find that
it is often easier and more computationally efficient to compute the
integral of the intensity over the
\emph{simplest} region -- meaning the one with the fewest boundaries --
and combine it with the formalism from \S\ref{sec:solution-no-occ} to
compute the visible flux. These integrals may be over the unocculted dayside,
the occulted dayside, the unocculted nightside, or the occulted nightside;
in the case of the latter three, a bit of algebra
(\S\ref{sec:cases-hard}--\S\ref{sec:cases-pathological}) is needed
to relate these to the observed flux. Additionally, in some cases we can
avoid computing new integrals entirely, as the solution can be obtained from
a combination of the classical \starry solution vector $\sT$ and the
formalism from the unocculted case (\S\ref{sec:solution-no-occ}).

@@

Cases with zero roots (case 1 -- case 5) are treated in \S\ref{sec:cases-easy},
while cases with one or two roots (case 6 -- case 10) are treated in
\S\ref{sec:cases-hard}. Cases with three or four roots
(case 11 -- case 14) are rarely encountered
in practice, but are possible for some pathological configurations; these are
treated in \S\ref{sec:cases-pathological}.

@@

\subsection{The term $\mathbb{p}_2$}
\label{sec:p2}
%
The final integral we must solve is that corresponding to
$\mathbb{p}_2$ ($\mu = \nu = 1$). As in \citet{Luger2019}, this is
the integral of the linear limb darkening term,
whose solution must be handled
separately due to the fact that the corresponding antiderivative
in Equation~(\ref{eq:G}) is not a polynomial in $x$, $y$, and $z(x, y)$;
also see \citet{Agol2020}.
The integral we must solve is
%
\begin{align}
    \label{eq:p2}
    \mathbb{p}_2 & =
    \vint{\pmb{\phi}}{
        \frac{1}{3}
        \left(
        \frac{
            1 - z(r_o\cos\varphi, b_o + r_o\sin\varphi)^3
        }{
            1 - z(r_o\cos\varphi, b_o + r_o\sin\varphi)^2
        }
        \right)
        \left(r_o^2 + b_o r_o \sin\varphi\right)
    }
    \quad,
\end{align}
%
where $z$ is the usual Cartesian coordinate (Equation~\ref{eq:z}).
The solution is tricky, but fortunately a similar integral was solved in
Equation~(34) of \citet{Pal2012}. Adapting their solution to our formalism,
we obtain
%
\begin{proof}{p2}
    \label{eq:p2_soln}
    \mathbb{p}_2 & =
    \frac{1}{3}
    \Big(
    c_0 +
    c_1 \DF +
    c_2 \DE +
    c_3 \DPi
    \Big)
\end{proof}
%
where
%
\begin{proof}{p2}
    c_0 &=
    \Delta
    \Bigg\{
    -\atantwo\left(
    -(b_o - r_o) \cos\valpha, (b_o + r_o) \sin\valpha
    \right)
    + \valpha
    \nonumber \\
    &\qquad\enspace\enspace
    - \frac{4}{3} b_o r_o
    \sin\valpha \cos\valpha
    \sqrt{1 - (b_o - r_o)^2 - 4 b_o r_o \sin^2\alpha}
    \nonumber \\
    &\qquad\enspace\enspace
    + \pmb{\delta}(b_o, r_o, \valpha)
    \Bigg\}
    \nonumber \\[0.5em]
    c_1 &=
    \frac{1 + b_o^4 - b_o^2(5 + 2 r_o^2) + r_o^4 + r_o^2}{3 \sqrt{1 - (b_o - r_o)^2}}
    \nonumber \\[0.5em]
    c_2 &= \frac{(b_o^2+7r_o^2-4)\sqrt{1 - (b_o - r_o)^2}}{3}
    \nonumber \\[0.5em]
    c_3 &= \frac{b_o + r_o}{(b_o - r_o)\sqrt{1 - (b_o - r_o)^2}}
    \quad,
    \\
    \intertext{and}
    %
    \delta(b_o, r_o, \alpha) &=
    \begin{cases}
        -2\pi
         &
        \qquad
        \alpha > \frac{3\pi}{2} \,\, \text{and} \,\, b_o > r_o
        \\
        +2\pi
         &
        \qquad
        \alpha > \frac{3\pi}{2} \,\, \text{and} \,\, b_o < r_o
        \\
        0
         &
        \qquad
        \text{otherwise}
        \quad.
    \end{cases}
\end{proof}
%
The quantities $\pmb{F}(\valpha \,|\, \nicefrac{1}{k^2})$ and
$\pmb{E}(\valpha \,|\, \nicefrac{1}{k^2})$
are the same incomplete elliptic integrals as those in
\S\ref{sec:J}, while
%
\begin{align}
    \label{eq:Pi}
    \Pi(n; \psi \,|\, m) & \equiv
    \int_0^{\psi}
    \frac{\dd \varphi}{(1 - n \sin^2\varphi)\sqrt{1 - m \sin^2 \varphi}}
\end{align}
%
is the incomplete elliptic integral of the third
kind, with
%
\begin{proof}{p2}
    \label{eq:n}
    n & = -\frac{4 b_o r_o}{(r_o - b_o)^2}
    \quad.
\end{proof}
%
